---
title: "Bay Area Bike Share: a Reference for R in Data Analytics"
author: "Adam Benson, PhD"
date: "9/18/2020"
output: html_document
---

Data taken from the Kaggle challenge 
"https://www.kaggle.com/benhamner/sf-bay-area-bike-share"

This data set has been used as a 'Hello World' for data science method 
development. After the data has been loaded, cleaned and is generally 
understood, it is customary to follow these steps:

1. Define the problem or stated objectives
2. Define or identify the desired metrics of results
3. Understand how to present the expected results 
4. Select candidate methods of analysis 
5. Identify data required to perform the analysis
6. Organize a method to gather data
7. Perform analysis and capture deviations from expected results 

and more...

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Packages <- c("readr", "data.table", "lubridate", "dplyr", "ggplot2") 
lapply(Packages, library, character.only = TRUE)
```


# Time-Based Exploration of Bicycle Trip Data

In this series of plots we will explore some data from the Bay Area Bicycle 
Share program. The purpose is to determine if we can identify patters from what 
might be thought of as noisy data sets. This is important given KR will be 
collecting, organizing, analyzing, and reporting using novel data sets.

Additional data exploration will help us develop a better research question, 
e.g. can we get a sense for how the frequency of use varies across time. Is the 
number of bicycle trips increasing or decreasing across time? Are there patterns
in the total number of bicycle trips according to the time of day or time of 
year? In this series of plots we will begin to explore these questions.

```{r}
##Load trip data
trip <- read.csv("trip.csv")
station <- read.csv("station.csv")
# station_temp_count <- station_temp %>% group_by(city) %>% count()

#Prepare data for analysis 
    #Get dates in right format
    trip$start_date <- mdy_hm(trip$start_date)
    trip$end_date <- mdy_hm(trip$end_date)
    
    trip$date <- trip$start_date
    trip$date <- as.Date(trip$date) 

    #Based on the start_station_temp ID, merge the "city" variable into trip df
    trip$date <- as.Date(trip$start_date)
    trip$id2 <- trip$id
    trip$id <- trip$start_station_id 
    trip <- left_join(trip, station, by = c ("start_station_id"="id"))
    
#Variable list
names(trip)
```

We see the data columns that are of interest to us.

## Trips by calendar date

First let's look at the total number of trips by calendar date. It will be 
interesting to see how the number of total trips varies throughout the year. 
Perhaps the number of trips on summer days will be more than the number of trips
during the winter. Let's find out.

```{r}
#Trips by calendar date 
datefreq <- count(trip, date)

ggplot(data = datefreq, aes(date, n)) +
    geom_point() +
    geom_smooth() +
    ggtitle("Trips Each Day") +
    ylab("Total Number of Bicycle Trips") +
    xlab("Date")
```
This plot shows the total number of trips each day for a two year period, from 
August 2013 through August 2015.

Looking at the smoother line, there appear to be fewer trips in January of 2014 
(on average) than by July of that year. While the number of trips on average 
appears to decrease until January of 2015, it doesn't appear to return to the 
same low of January 2014. From January 2015 until July 2015 the average number 
of daily trips appears to be increasing. However, we can see an apparent 
seasonal adjustment in rental activity.

Also, there appears to be an interesting split in the data. Perhaps there is a 
confounding variable. My guess is that is has to do with what day of the week it
is. For example, do more people make bike trips on the weekends when they have 
more leisure time? Or maybe the commuters are taking the bikes to get to work.
Perhaps Friday is a big day for biking as it is the end of the work week? Let's 
find out!

## Total number of trips by day of the week

```{r}
dailyfreq <- as.data.frame(table(wday(trip$date, label = TRUE)))
    
ggplot(data = dailyfreq, aes(x = Var1, Freq)) +
    geom_bar(stat="identity") + #need to include stat = identity or will 
    #try to make a count of the count 
    ggtitle("Total Number of Trips Per Day") +
    ylab("Total Number of Bicycle Trips") +
    xlab("Day of the Week")
```

It appears the commuter hypotHesis holds. It appears that fewer trips occur on 
the weekends than on the weekdays. Commuters dominate the usage of the bicycles 
which may explain this pattern. That's something to consider in future analysis 
(e.g., binary variable). We also need to make sure that this pattern explains 
the division in the data we saw earlier. This information could help the company
management be better prepared for the scheduling of the relocation staff

In order to see how the data are split, code the prior calendar plot with colors
that correspond with weekdays or weekends. Remember that packaging sells.

## Total number of trips by calendar date - weekend vs. weekday

```{r}
datefreq <- mutate(datefreq, weekend = (wday(datefreq$date) == 1 |
                    wday(datefreq$date) == 7))
            #Makes variable with True if date == sunday(1) or saturday (7)
        
datefreq$weekend <- factor(datefreq$weekend, labels = c("Weekday", "Weekend"))
    #Give labels to variable

ggplot(data = datefreq, aes(date, n)) +
    geom_point(aes(color = weekend), size = 3, alpha = 0.65) +
    ggtitle("Total Number of Trips Per Day") +
    ylab("Total Number of Bicycle Trips") +
    xlab("Date")

```

Interesting! This confirms that some of the pattern in the data across the year 
can be explained by weekday vs. weekend usage. Now get a view of the same data 
but plotted separately by weekend and weekday. Note you see that during the 
holiday season, even weekly rental seems to drop off.

Separate plots for weekend and weekday

```{r}
ggplot(data = datefreq, aes(date, n)) +
            geom_point(size = 3, alpha = 0.65) +
            facet_grid(. ~ weekend) + 
            geom_smooth(se = FALSE) +
            ylab("Total Number of Bicycle Trips") +
            xlab("Date")
```
With the data split one can see a little better the trends between weekday and 
weekend usage over time. Weekday trips still appear to have a lot of variance. 
Cold this variance be explained by the holidays. look at the end of December and
beginning of January.

Now that we've seen how the number of trips varies throughout the year, how 
about we take a look at how it varies throughout the day.

## Total trips by hour of the day

Given the majority of the rental happens during week days, commuting days, it 
would follow that rental activity would mirror hourly commuting patterns.

```{r}
# Note: Since R doesn't have a time only class, you can convert the time stored 
# in ymd_hms (via lubridate) to hours and minutes in numeric format 

t2 <- ymd_hms(trip$start_date) 
t3 <- hour(t2) + minute(t2)/60
trip$daytime <- t3 
rm(t2, t3) #Cleanup 
    
ggplot(trip, aes(daytime)) +
  geom_histogram(binwidth = 0.25) + #Every fifteen minutes = binwidth 
  geom_vline(xintercept = 9, color = 'orange')+
  geom_vline(xintercept = 17, color = 'red', alpha = 0.7) +
  annotate("text", x = 9, y = 27000, label = "9:00 AM", color = "orange",
           size = 7) +
  annotate("text", x = 17, y = 27000, label = "5:00 PM", color = "red", 
           size = 7) +
  xlab("Time of day on 24 hour clock") +
  ylab("Total number of bicycle trips")
```
The rental patterns look to line up well with what we might consider 
traditional commuting patterns.

## Number of trips by hour, across the year


```{r}
trip$quarter <- quarter(trip$date)

ggplot(trip, aes(daytime)) +
  geom_histogram(binwidth = 0.25) + #Every fifteen minutes = binwidth 
  geom_vline(xintercept = 9, color = 'orange')+
  geom_vline(xintercept = 17, color = 'red', alpha = 0.7) +
  xlab("Time of day on 24 hour clock") +
  ylab("Total number of bicycle trips") +
  facet_wrap(~quarter)
```
This breaks the rental data out across the quarter (season) showing that there 
is a lot of activity during the commuting time as well as at lunch.

Each plot corresponds with a different quarter of the year. Quarter 1 (January -
March), Quarter 2 (April - June), Quarter 3 (July - September) or Quarter 4 
(October - December). From these plots it looks like that pattern from before 
holds. That is, the total number of trips peaks around rush hour each. Also 
notice the consistent small peak around lunch hour each day at noon.

Now let's take a look at how the city and type of bicycle rider (subscriber vs. 
customer) may be influencing these trends.

## Usage by city

```{r}
#Weekend variable to trip df
trip <- mutate(trip, weekend = (wday(trip$date) == 1 | wday(trip$date) == 7))
trip$weekend <- factor(trip$weekend, labels = c("Weekday", "Weekend"))
    
#Plot usage by city
ggplot(data = trip, aes(date)) +
  geom_bar(aes(color = weekend), stat = "count",
              position = "stack") +
    ggtitle("Trips by City Across Time") +
    ylab("Total Number of Bicycle Trips") +
    xlab("Trend Across Time") +
    facet_grid(~city) +
    theme(axis.text.x = element_blank())
```

As you can see, San Francisco dominates the usage of the program. One thing one 
should also consider is if customers vs. subscriber is influencing anything.
Understanding this could again help with staffing, but marketing and plans for 
expansion as well.

## Customers vs. Subscribers

```{r}
ggplot(data = trip, aes(date)) +
  geom_bar(aes(color = subscription_type), stat = "count", 
           position = "stack") +
  ggtitle("Customer Vs. Subscriber on Weekends and Weekdays") +
  ylab("Total Number of Bicycle Trips") +
  xlab("Trend Across Time") +
  facet_grid(~weekend) +
  theme(axis.text.x = element_blank())
```

It looks like subscribers dominate usage on the weekday. Weekend the usage is 
more balanced. Does the strength of this trend hold for different cities? 
Perhaps with all those tourists there are more customers in San Francisco 
relative to subscribers.

```{r}
ggplot(data = trip, aes(date)) +
  geom_bar(aes(color = subscription_type), stat = "count", position = "stack") +
  ggtitle("Subscribers Vs. Customers - Trips Per Day by City ") +
  ylab("Total Number of Bicycle Trips") +
  xlab("Trend Across Time") +
  facet_wrap(~city, scale = "free_y") +
  theme(axis.text.x = element_blank())
```

It looks like the trend does indeed hold for San Francisco. However, Palo Alto 
seems to have a more balanced usage. You can also get a sense from these graphs 
how unbalanced the usage is across cities. My hometown of Redwood City peaks at 
about 25 trips a day compared with San Francisco that peaks closer to 1,300 
trips a day.

We've seen how the number of trips fluctuates across the entire year, how it 
fluctuates according to weekend Vs. weekday, and by hour of the day. We've also 
seen the balance of the number of trips by city and by subscription type.

What would you suggest the next research questions be?
Can you put a model to the data to predict usage? The data is non-linear...
Given the data sets we expect at KR what strategy cold help develop the most 
pertinent information?

