----
title: "Bay Area Bike Share: a Reference for R in Data Analytics - Machine Learning"
author: "Adam Benson, PhD"
date: "9/29/2020"
output: html_notebook
---

Data taken from the Kaggle challenge 
"https://www.kaggle.com/benhamner/sf-bay-area-bike-share"

This data set has been used as a 'Hello World' for data science method 
development. After the data has been loaded, cleaned and is generally 
understood, it is customary to follow these steps:

1. Define the problem or stated objectives
2. Define or identify the desired metrics of results
3. Understand how to present the expected results 
4. Select candidate methods of analysis 
5. Identify data required to perform the analysis
6. Organize a method to gather data
7. Perform analysis and capture deviations from expected results 

and more...

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Packages <- c("readr", "data.table", "lubridate", "dplyr", "ggplot2") 
lapply(Packages, library, character.only = TRUE)
```
## R Markdown
```{r}
## Load trip data

true_trip <- read.csv("trip.csv")
true_station <- read.csv("station.csv")
```


```{r}
# Do this so you don't need to read the csv in if you want to start over.
trip <- true_trip
station <-true_station

# Prepare data for analysis 
# Get dates in POSIXct, format
trip$start_date <- mdy_hm(trip$start_date)
trip$end_date <- mdy_hm(trip$end_date)
str(trip)
```

```{r}
trip$date <- trip$start_date
trip$date <- as.Date(trip$date) 
#Based on the start_station_temp ID, merge the "city" variable into trip df
trip$date <- as.Date(trip$start_date)
trip$end_station_name <- as.factor(trip$end_station_name)

trip$zip_code <- as.factor(trip$zip_code)
trip$id2 <- trip$id
trip$id <- trip$start_station_id
trip <- left_join(trip, station, by = c ("start_station_id"="id"))
str(trip)
```


```{r}
trip$city <- as.factor(trip$city)
trip$installation_date <- mdy(trip$installation_date)

trip$start_station_id <- as.factor(trip$start_station_id)
trip$start_station_name <- as.factor(trip$start_station_name)

trip$end_station_id <- as.factor(trip$end_station_id)
trip$end_station_name <- as.factor(trip$end_station_name)

trip$id <- as.factor(trip$id)
trip$id2 <- as.factor(trip$id2)

trip$bike_id <- as.factor(trip$bike_id)

trip$subscription_type <- as.factor(trip$subscription_type)

trip$name <- as.factor(trip$name)
trip$city <- as.factor(trip$city)

summary(trip)
```


```{r}
str(trip)
```
Verify that the data is usable

```{r}
#Trips by calendar date 
datefreq <- count(trip, date)

ggplot(data = datefreq, aes(date, n)) +
    geom_point() +
    geom_smooth(span = .3) + # (span = .7) is about the the default, gray lines represent SE
    ggtitle("Trips Each Day") +
    ylab("Total Number of Bicycle Trips") +
    xlab("Date")
```

```{r}
datefreq <- mutate(datefreq, weekend = (wday(datefreq$date) == 1 |
                    wday(datefreq$date) == 7))
            #Makes variable with True if date == sunday(1) or saturday (7)
        
datefreq$weekend <- factor(datefreq$weekend, labels = c("Weekday", "Weekend"))
    #Give labels to variable

ggplot(data = datefreq, aes(date, n)) +
    geom_point(aes(color = weekend), size = 3, alpha = 0.65) +
    ggtitle("Total Number of Trips Per Day") +
    ylab("Total Number of Bicycle Trips") +
    xlab("Date")

```

## Total number of trips by day of the week

```{r}
dailyfreq <- as.data.frame(table(wday(trip$date, label = TRUE)))
    
ggplot(data = dailyfreq, aes(x = Var1, Freq)) +
    geom_bar(stat="identity") + #need to include stat = identity or will 
    #try to make a count of the count 
    ggtitle("Total Number of Trips Per Day") +
    ylab("Total Number of Bicycle Trips") +
    xlab("Day of the Week")
```



```{r}
# Note: Since R doesn't have a time only class, you can convert the time stored 
# in ymd_hms (via lubridate) to hours and minutes in numeric format 

t2 <- ymd_hms(trip$start_date) 
t3 <- hour(t2) + minute(t2)/60
trip$daytime <- t3 
rm(t2, t3) #Cleanup 
    
ggplot(trip, aes(daytime)) +
  geom_histogram(binwidth = 0.25) + #Every fifteen minutes = binwidth 
  geom_vline(xintercept = 9, color = 'orange')+
  geom_vline(xintercept = 17, color = 'red', alpha = 0.7) +
  annotate("text", x = 9, y = 27000, label = "9:00 AM", color = "orange",
           size = 7) +
  annotate("text", x = 17, y = 27000, label = "5:00 PM", color = "red", 
           size = 7) +
  xlab("Time of day on 24 hour clock") +
  ylab("Total number of bicycle trips")
```
```{r}

head(datefreq,10)
```

