---
title: "Bay Area Bike Share: a Reference for R in Data Analytics"
author: "Adam Benson, PhD"
date: "7/22/2020"
output: html_document
---

Data taken from the Kaggle challenge 
"https://www.kaggle.com/benhamner/sf-bay-area-bike-share"

This data set has been used as a 'Hello World' for data science method 
development. After the data has been loaded, cleaned and is generally 
understood, it is customary to follow these steps:

1. Define the problem or stated objectives
2. Define or identify the desired metrics of results
3. Understand how to present the expected results 
4. Select candidate methods of analysis 
5. Identify data required to perform the analysis
6. Organize a method to gather data
7. Perform analysis and capture deviations from expected results 

and more...

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
Packages <- c("readr", "data.table", "lubridate", "dplyr", "ggplot2") 
lapply(Packages, library, character.only = TRUE)
```

# Extract, Transform, and Load (ETL) Bike Share Data

The plan is to load each data file, perform superficial data exploration and 
clean up before moving forward. In the cases where the data needs to 

## Station Data

Example of loading a csv into R

station.csv - Contains data that represents a station where users can pickup or return bikes.

```{r get_data_station}
station <- data.table(read.csv("~/Code/solo_projects/bike_share/station.csv", 
                               stringsAsFactors=TRUE))#
```

Quick check

```{r get_data_station_review}
str(station) # 
```

id is shown as an int and installation is listed as a factor. Fix this. 
Everything else looks good - unless there are some missing values.

Manipulate the data to be the correct form.

```{r get_data_station_reformat}
station$id <- as.factor(station$id)
station$installation_date <- mdy(station$installation_date)
summary(station)
```

The data in this table now looks reasonable.

# Trip

trips.csv - Data about individual bike trips

This provides a trip ID, duration of trip in minutes, the DTS of the start of 
the rental, a couple of identifying factors of the station where the trip began, 
a couple of identifying factors of the station where the trip ended, the ID of 
the bike that was rented, whether the person renting did so as part of a 
subscription or one time event, the zip code is of the renter.

```{r get_data_trip}
trip <- data.table(read.csv("~/Code/solo_projects/bike_share/trip.csv", stringsAsFactors=TRUE))
str(trip)
```

ID is listed as a number, dates are listed as factors, station ids are listed as
numbers, bike ids are listed as numbers. These issues should be fixed.

```{r get_data_trip_correction}
trip$id <- as.factor(trip$id)
trip$start_date <- mdy_hms(trip$start_date)
trip$end_date <- mdy_hms(trip$end_date)
trip$start_station_id <- as.factor(trip$start_station_id)
trip$end_station_id <- as.factor(trip$end_station_id)
trip$bike_id <- as.factor(trip$bike_id)
summary(trip)
```

Trip looks to be better. But look at the max value for duration. For the purpose
of this exercise, we will remove that record. Normally you do not remove data, 
ever. In this case, it is a report that the bike has been out for more than 32 
years - not possible. We will look at rentals that are a year or less.

```{r}
new_trip <- subset(trip, duration <= 525600 )
summary(new_trip)
```


## Weather

weather.csv - Data about the weather on a specific day for certain zip codes

```{r get_data_weather}
weather <- data.table(read.csv("~/Code/solo_projects/bike_share/weather.csv", 
                               na.strings = c("")))
str(weather) # 
```

Date, precipitation_inches, and zip are wrong. Events should be cleaned up.

```{r get_data_weather_corrections}
weather$date <- mdy(weather$date)
summary(weather)
weather$zip_code <- as.factor(weather$zip_code)
summary(weather)
weather$precipitation_inches <- 
  as.numeric(replace(weather$precipitation_inches, 
                     weather$precipitation_inches == 'T', 0))
summary(weather)
weather$events <- as.factor(tolower(weather$events))
str(weather)
```

Weather looks a good deal better, but there are still some issues. 
'wind_dir_degrees' has a max value of 2772 and there is no such value in 
degrees. how should this be handled?
Look at 'max_gust_speed_mph" and 'max_wind_Speed_mph' The values are 114.0 and 
128 respectively. is that correct? That wind speed seems unlikely in the Bay 
Area.

## Status

status.csv - data about the number of bikes and docks available for given station and minute.

This is a larger data file, with almost 72 million records. data.table will help a lot here. Spoiler alert: station id and time are categorized as the wrong type of data. Let's Save the suspense and fix them now.

```{r get_data_status}
status <- data.table(read.csv("~/Code/solo_projects/bike_share/status.csv"))
status$station_id <- as.factor(status$station_id)
status$time <- ymd_hms(status$time)
str(status)
```

Next, we can propose any number of analyses, then select one to perform [next meeting].

## Visualize the Data

### Stations

Number of stations in each city

```{r}
station_count <- station %>% group_by(city) %>% count()

setnames(station_count, "n", "stations")

ggplot(station_count, aes(x=city, y=stations, fill=stations))+
  geom_bar(stat="identity") +
  geom_text(aes(label=stations), vjust=-0.3, size=4)+
  xlab("City") + ylab("Station Count") 
```

What does this chart show/tell you?


### Trips

How long are bikes typically rented?

Maybe a boxplot could help...
```{r}
new_trip %>% ggplot(aes(x=subscription_type, y=duration/60/24, fill=subscription_type)) +
  geom_boxplot() +
  geom_jitter(color="black", size=0.4, alpha=0.9) +
  theme(
    legend.position="none",
    plot.title = element_text(size=11)
  ) +
  ggtitle("A boxplot of rentals(a year or less)") +
  xlab("") + ylab("days rented")

```

It look like there are so many data points, with some outliers, the boxplot is nearly useless.

A density plot will show the term of rental as a percentage of the whole.

```{r}
month_or_less <- subset(trip, duration <= 60*24*7) # 7 days
ggplot(month_or_less, aes(fill = subscription_type, x = (duration/60/24)))+
  geom_density() +
  scale_fill_manual(values=c("#0000ff", "#ff8800")) + 
  # theme_classic() +
  xlab("days rented") +
  ylab("percent of business")
```

 This right tailed graph shows that there are very few rentals more than two days long.

```{r}
day_or_less <- subset(trip, duration <= 60*24) # 1 day
ggplot(day_or_less, aes(fill = subscription_type, x = (duration/60)))+
  geom_density() +
  scale_fill_manual(values=c("#0000ff", "#ff8800")) + 
  # theme_classic() +
  xlab("hours rented") +
  ylab("percent of business") +
  ggtitle("hours rented by type of customer")

```

The rental business might want to know more about day-to-day rentals for inventory management.

### Weather

Maybe weather, temperature, would have an effect on rentals. Look for extreme values in temperature before looking for a correlation.

```{r}
summary(weather$max_temperature_f)
```

```{r}
summary(weather$min_temperature_f)
```
```{r}
boxplot(weather$min_temperature_f~cycle(weather$mean_temperature_f))
```

```{r}
plot(weather$date,weather$max_temperature_f)
```

```{r}
ggplot(weather, aes(x=date, y=mean_temperature_f)) +
  geom_line()

```

As you might expect, there is a cyclical pattern to the temperatures. Average 
max temp was mid 80'2 and average min temperature was the high 30's. Relatively 
speaking, the Bay Area weather is fairly mild. Winters are typically cool and 
wet; summers are warm and dry. Later we can look for a model fit of weather 
explaining variance in bike rental.

### Status

We might look at how many bikes are rented out as the docks available minus the 
bikes available we don't know the number of bikes in the total inventory at a
given time, but if docks are added, by way of bringing on new stations, 
it might be safe to guess the number of bikes have been added to fill those 
docks.
```{r}
summary(status)
```







